Periodograms for Multiband Astronomical Time Series

### 2015-05-15 writeup/paper.tex
multiplicative -> harmonic

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{fig07a.pdf}
  \includegraphics[width=\textwidth]{fig07b.pdf}
  \caption{
    Comparison of the Multiband algorithm and single-band supersmoother algorithm on 483 well-sampled RR Lyrae light curves from Stripe 82.
    The upper panels show a representative lightcurve and periodogram fits, while the bottom panels compare the derived periods to the template-based periods reported in S10.
    Shown for reference are the beat aliases (dotted lines) and the multiplicative alias (dashed lines): numbers along the top and right edges of the panels indicate the number of points aligned with each trend.
    The single-band supersmoother model tends to err toward multiplicative aliases, while the multiband model tends to err toward beat frequency aliases.
    Both methods find the correct period among the top 5 significant peaks around 99\% of the time.
    \new{This suggests that for densely-sampled multiband surveys, the multiband periodogram will match the results of standard methods (but see \fig{compare_periods_reduced}).}
  } 
  \figlabel{compare_periods}
\end{figure*}

@@

The lower panels of \fig{compare_periods} compare the S10 period with the best periods obtained from the 1-band supersmoother (lower-left) and from the shared-phase multiband model (lower-right). To guide the eye, the figure includes indicators of the locations of beat aliases (dotted lines) and multiplicative aliases (dashed lines) of the S10 period. \new{Numerical results are summarized in the upper rows of Table \ref{table:results}.}

The best-fit supersmoother period matches the S10 period in 87\% of cases (421/483), while the best-fit multiband period matches the S10 period in 79\% of cases (382/483). The modes of failure are instructive: when the supersmoother model misses the S10 period, it tends to land on a multiplicative alias (i.e. the dashed line). This is due to the flexibility of supersmoother: a doubled period spreads the points out, leading to fewer constraints in each neighborhood and thus a smaller average residual around model. In other words, the SuperSmoother tends to over-fit data which is sparsely-sampled. On the other hand, when the multiband model misses the S10 period, it tends to land on a beat alias between the S10 period and the 1-day observing cadence (i.e. the dotted lines). This is due to the fact that the single-frequency periodic model is biased, and significantly under-fits the data: it cannot distinguish residuals due to underfitting from residuals due to window function effects.

@@

\begin{table*}
\centering
  \caption{Period Determination from dense and sparse data (483 total)}
  \begin{tabular}{|l|l|l|l|l|l|l|}
  \hline
   Data & Method & Match among top 5 & Top peak matches & Beat Aliases & Multiplicative Aliases  \\
  \hline\hline
  Dense data (\fig{compare_periods})
  & g-band Supersmoother & 477 (98.8\%) & 421 (87.2\%)& 31 & 34 \\
  & Multi-band Periodogram & 480 (99.4\%) & 382 (79.1\%) & 94 & 5 \\
  \hline\hline
  Sparse data (\fig{compare_periods_reduced})
  & g-band Supersmoother & 219 (45.3\%) & 113 (23.4\%) & 101 & 4 \\
  & Multi-band Periodogram & 449 (93.0\%) & 308 (63.8\%) & 136 & 7 \\
  \hline
  \end{tabular}
  \label{table:results}
\end{table*}

>>>>

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{fig07a.pdf}
  \includegraphics[width=\textwidth]{fig07b.pdf}
  \caption{
    Comparison of the Multiband algorithm and single-band supersmoother algorithm on 483 well-sampled RR Lyrae light curves from Stripe 82.
    The upper panels show a representative lightcurve and periodogram fits, while the bottom panels compare the derived periods to the template-based periods reported in S10.
    Shown for reference are the beat aliases (dotted lines) and the \old{multiplicative} \new{first harmonic} alias (dashed lines): numbers along the top and right edges of the panels indicate the number of points aligned with each trend.
    The single-band supersmoother model tends to err toward \old{multiplicative} \new{harmonic} aliases, while the multiband model tends to err toward beat frequency aliases.
    Both methods find the correct period among the top 5 significant peaks around 99\% of the time.
    \new{This suggests that for densely-sampled multiband surveys, the multiband periodogram will match the results of standard methods (but see \fig{compare_periods_reduced}).}
  } 
  \figlabel{compare_periods}
\end{figure*}

@@

The lower panels of \fig{compare_periods} compare the S10 period with the best periods obtained from the 1-band supersmoother (lower-left) and from the shared-phase multiband model (lower-right). To guide the eye, the figure includes indicators of the locations of beat aliases (dotted lines) and \old{multiplicative} \new{first harmonic} aliases (dashed lines) of the S10 period. \new{Numerical results are summarized in the upper rows of Table \ref{table:results}.}

The best-fit supersmoother period matches the S10 period in 87\% of cases (421/483), while the best-fit multiband period matches the S10 period in 79\% of cases (382/483). The modes of failure are instructive: when the supersmoother model misses the S10 period, it tends to land on a \old{multiplicative} \new{harmonic} alias (i.e. the dashed line). This is due to the flexibility of supersmoother: a doubled period spreads the points out, leading to fewer constraints in each neighborhood and thus a smaller average residual around model. In other words, the SuperSmoother tends to over-fit data which is sparsely-sampled. On the other hand, when the multiband model misses the S10 period, it tends to land on a beat alias between the S10 period and the 1-day observing cadence (i.e. the dotted lines). This is due to the fact that the single-frequency periodic model is biased, and significantly under-fits the data: it cannot distinguish residuals due to underfitting from residuals due to window function effects.

@@

\begin{table*}
\centering
  \caption{Period Determination from dense and sparse data (483 total)}
  \begin{tabular}{|l|l|l|l|l|l|l|}
  \hline
   Data & Method & Match among top 5 & Top peak matches & Beat Aliases & \old{Multiplicative} \new{Harmonic} Aliases  \\
  \hline\hline
  Dense data (\fig{compare_periods})
  & g-band Supersmoother & 477 (98.8\%) & 421 (87.2\%)& 31 & 34 \\
  & Multi-band Periodogram & 480 (99.4\%) & 382 (79.1\%) & 94 & 5 \\
  \hline\hline
  Sparse data (\fig{compare_periods_reduced})
  & g-band Supersmoother & 219 (45.3\%) & 113 (23.4\%) & 101 & 4 \\
  & Multi-band Periodogram & 449 (93.0\%) & 308 (63.8\%) & 136 & 7 \\
  \hline
  \end{tabular}
  \label{table:results}
\end{table*}
